---
title: "Adam_Section_Go_Brrr"
output: html_document
date: "2025-04-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(ggplot2)
library(scales)
library(plotly)

library(readr)
diabetes_dataset <- read_csv("diabetes_prediction_dataset.csv")
#View(diabetes_dataset)
```


## Smokers go brrr

In the smoking data there are 6 unique values

1)  Never: Has Never smoked
2)  Not current: Has smoked but is not currently smoking
3)  Former: Has quit smoking (abstained for longer than)
4)  Current: Is currently a smoker
5)  Ever: Has ever smoked regardless of current smoking status
6)  No Info: No smoking history information available

The total amount of people who fall into each category is as follows;

1)  Never: 35095
2)  Not current: 6447
3)  Former: 9352
4)  Current: 9286
5)  Ever: 4004
6)  No Info: 35816

There is quite a sizable amount of people in the No info category.

The total number of people in the dataset is `r nrow(diabetes_dataset)`. To help clean up the data, we can filter 'No Info' people out. When we do that we get `r nrow(diabetes_dataset %>% filter(smoking_history != 'No Info'))`.

```{r Cleaning smoking data}

# Figure out the unique categories of smoking history
unique(diabetes_dataset$smoking_history)

# Count amount of people who belong to each unique smoking category
#Omit No info

smoking_tally <- diabetes_dataset %>% filter(smoking_history != 'No Info') %>%  group_by(smoking_history) %>% summarise(total_people = n())

#group diabetic vs non diabetic people together

smoking_diabetes_dataset <- diabetes_dataset %>%
  filter(smoking_history != 'No Info') %>%  
  group_by(smoking_history, diabetes) %>%  
  summarise(total = n())

smoking_diabetes_dataset

# Inner join tally data with diabetic grouped data;
#mutate a column to create a percentage per category;
#select desired columns

smoking_diabetes_percentage <- inner_join(smoking_tally, smoking_diabetes_dataset, by = 'smoking_history') %>% mutate(Percentage = round( total/total_people, digits = 4)) %>% select(smoking_history, diabetes, total, Percentage)

smoking_diabetes_percentage 
```

Now we can graph the relationship between smoking and diabetes as separated by smoking category.

```{r Plotting smoking data}
library(ggplot2)
library(plotly)

#Create initial graph about smoking and diabetes

smoking_diabetes_graph <- ggplot(smoking_diabetes_percentage,
       aes(x    = smoking_history,
           y    = total,
           fill = factor(diabetes))) +
  geom_col(position = position_dodge(width = 0.9)) +
 
  # add percent labels
  geom_text(aes(label = percent(Percentage, accuracy = .1)),
            position = position_dodge(width = 1),
            vjust    = .4,        # nudge above bar
            size     = 3,
            hjust    = -.1, size = 3) +
            
 
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, .1))) +
  labs(
    title   = "Smoking History and Diabetes Relationship",
    x       = "Smoking History",
    y       = "Count of People",
    fill    = "Has Diabetes",
    caption = "1 = diabetes, 0 = no diabetes"
  ) 

smoking_diabetes_graph
```




```{r heart disease and smoking}
smoking_heart_disease_data <- diabetes_dataset %>%
  filter(smoking_history != 'No Info') %>%  
  group_by(smoking_history, heart_disease) %>%  
  summarise(total = n())
smoking_heart_disease_data

smoking_heart_disease_percentage <- inner_join(smoking_tally, smoking_heart_disease_data, by = 'smoking_history') %>% mutate(Percentage = round( total/total_people, digits = 4)) %>% select(smoking_history, heart_disease, total, Percentage)

smoking_heart_disease_percentage

```




```{r}
ggplot(smoking_heart_disease_percentage,
       aes(x    = smoking_history,
           y    = total,
           fill = factor(heart_disease))) +
  geom_col(position = position_dodge(width = 0.9)) +
 
  # add percent labels
  geom_text(aes(label = percent(Percentage, accuracy = .1)),
            position = position_dodge(width = 1),
            vjust    = .4,        # nudge above bar
            size     = 3,
            hjust    = -.1, size = 3) +
            
 
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, .1))) +
  labs(
    title   = "Smoking History and Heart Disease Relationship",
    x       = "Smoking History",
    y       = "Count of People",
    fill    = "Has Heart Disease",
    caption = "1 = Heart Disease, 0 = No Heart Disease"
  ) 

```




```{r Smoking and hypertension}
smoking_hypertension_data <- diabetes_dataset %>%
  filter(smoking_history != 'No Info') %>%  
  group_by(smoking_history, hypertension) %>%  
  summarise(total = n())
smoking_hypertension_data

smoking_hypertension_percentage <- inner_join(smoking_tally, smoking_hypertension_data, by = 'smoking_history') %>% mutate(Percentage = round( total/total_people, digits = 4)) %>% select(smoking_history, hypertension, total, Percentage)

smoking_hypertension_percentage
```


```{r Hypertension Graph}
ggplot(smoking_hypertension_percentage,
       aes(x    = smoking_history,
           y    = total,
           fill = factor(hypertension))) +
  geom_col(position = position_dodge(width = 0.9)) +
 
  # add percent labels
  geom_text(aes(label = percent(Percentage, accuracy = .1)),
            position = position_dodge(width = 1),
            vjust    = .4,        # nudge above bar
            size     = 3,
            hjust    = -.1, size = 3) +
            
 
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, .1))) +
  labs(
    title   = "Smoking History and Hypertension Relationship",
    x       = "Smoking History",
    y       = "Count of People",
    fill    = "Has Hypertension",
    caption = "1 = Hypertension, 0 = No Hypertension"
  ) 

```




```{r}
library(scales)

# 2) Plot with geom_col + geom_text
ggplot(smoking_diabetes_percentage,
       aes(x    = smoking_history,
           y    = total,
           fill = factor(diabetes))) +
  geom_col(position = position_dodge(width = 0.9)) +
 
  # add percent labels
  geom_text(aes(label = percent(Percentage, accuracy = .1)),
            position = position_dodge(width = 1),
            vjust    = .4,        # nudge above bar
            size     = 3,
            hjust    = -.1, size = 3) +
            
 
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, .1))) +
  labs(
    title   = "Smoking History and Diabetes Relationship",
    x       = "Smoking History",
    y       = "Count of People",
    fill    = "Has Diabetes",
    caption = "1 = diabetes, 0 = no diabetes"
  ) 


```


