---
title: "Adam_Section_Go_Brrr"
output: ioslides_presentation
date: "2025-04-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

One of the more interesting bits of data is that there are individuals that have an a1c of over 6.5 yet are not considered diabetic.

```{r hbA1C vs Blood Sugar}

A1C_vs_Glucose <- diabetes_dataset %>%  ggplot(aes(x = blood_glucose_level , y = HbA1c_level, color = as.factor(diabetes))) + geom_point() + geom_jitter()

A1C_vs_Glucose


```


# Smokers go brrr

In the smoking data there are 6 unique values

1)  Never: Has Never smoked
2)  Not current: Has smoked but is not currently smoking
3)  Former: Has quit smoking (abstained for longer than)
4)  Current: Is currently a smoker
5)  Ever: Has ever smoked regardless of current smoking status
6)  No Info: No smoking history information available

The total amount of people who fall into each category is as follows;

1)  Never: 35095
2)  Not current: 6447
3)  Former: 9352
4)  Current: 9286
5)  Ever: 4004
6)  No Info: 35816

There is quite a sizable amount of people in the No info category.

The total number of people in the dataset is `r nrow(diabetes_dataset)`. To help clean up the data, we can filter 'No Info' people out. When we do that we get `r nrow(diabetes_dataset %>% filter(smoking_history != 'No Info'))`.

```{r Cleaning smoking data}

# Figure out the unique categories of smoking history
unique(diabetes_dataset$smoking_history)

# Count amount of people who belong to each unique category
diabetes_dataset %>% group_by(smoking_history) %>% summarise(total_people = n())

smoking_diabetes_dataset <- diabetes_dataset %>%
  filter(smoking_history != 'No Info') %>%  
  group_by(smoking_history, diabetes) %>%  
  summarise(total = n())
```

Now we can graph the relationship between

```{r Plotting smoking data}

print('hello world')

```

