---
title: "Final Project"
author: "Adam Guenoun, Indira Martinez, Nicholas Solis"
date: "`r Sys.Date()`"
output: html_document
---
### Introduction:

  Within this analysis, we'll investigate factors correlated to diabetes. With a data set of 100,000 people, this investigation allows us to display relations between ages, HbA1c levels, smoking history, and glucose levels. With a wide range of data points, we begin to question if there are trends within this data that match our general understanding of diabetes. Our goal is to asses which of the 9 variables play a stronger role to the development of diabetes and if we can prove trends to better support our assumptions of this data. Through data visualization, chart analysis, and numerical analysis we will be able to present this data to convicne a general audience of the important factors that contribute to diabteic trends. 





```{r, message=FALSE}

library(tidyverse) ## Loaded for dplyr
library(ggplot2) ## Loaded for plotting
library(plotly) ## Loaded for interactive plots
library(readr) ## Loaded to read in data
library(knitr) ## Loaded to compute and display data
library(scales) ## Loaded to scale data 
```

```{r Loading in Data, include=FALSE}
library(readr)
diabetes_dataset <- read_csv("diabetes_prediction_dataset.csv")
#View(diabetes_dataset)
```

### Diabetes Dataset

```{r, echo=FALSE,fig.width=4, fig.height=4 }
library(knitr)
#diabetes_dataset
kable(head(diabetes_dataset, 6), caption = "100,000 × 9 (first 6 rows)")
```



### Male vs. Female Blood Sugar Levels (HbA1c) Plot

```{r, echo=FALSE}
library(dplyr)
library(ggplot2)
library(plotly)
library(knitr) 

HbA1c_by_gender <- diabetes_dataset %>% filter(gender != 'Other') %>% 
  mutate(HbA1c_category = case_when(
    HbA1c_level < 5.7 ~ "Normal < 5.7%",
    HbA1c_level >= 5.7 & HbA1c_level < 6.5 ~ "Prediabetes 5.7% - 6.4%",
    HbA1c_level >= 6.5 ~ "Diabetes ≥ 6.5%",
    TRUE ~ NA_character_  # handles unexpected cases that don’t match the defined conditions keeping missing values as NA
    ))

HbA1c_by_gender <- HbA1c_by_gender %>%
  mutate(HbA1c_category = factor(HbA1c_category, levels = c("Normal < 5.7%", "Prediabetes 5.7% - 6.4%", "Diabetes ≥ 6.5%")))

# here we print out our mutated dataset to make a comparison of the original.
#HbA1c_by_gender
kable(head(HbA1c_by_gender, 6), caption = "99,982 × 10 (first 6 rows)")
```

Similar Prevalence of Prediabetes – The proportion of individuals categorized as having prediabetes (HbA1c 5.7% - 6.4%) is almost identical between males (41.3%) and females (41.4%). This suggests that prediabetes affects both genders at nearly the same rate.

-   Slightly Higher Diabetes Rates Among Males – More males (21.6%) fall into the diabetes (HbA1c ≥ 6.5%) category compared to females (20.2%). While the difference is small, it might indicate that men have a slightly higher risk of diabetes in this dataset.

**Females Have a Slightly Higher Proportion** of Normal Blood Sugar Levels – More females (38.4%) fall into the normal blood sugar category (\<5.7%) compared to males (37.1%). This may indicate some slight protective factors or lifestyle differences in this group.

Since more males are in the diabetes category, there could be gender-related risk factors worth exploring—such as diet, activity levels, or genetic predisposition.

Overall, blood sugar regulation patterns appear fairly balanced between genders, but small differences suggest potential areas for further investigation.

::: callout-note
**Similar Prevalence of Prediabetes**\
The proportion of individuals classified as having prediabetes (HbA1c 5.7% - 6.4%) is nearly identical between males (**41.3%**) and females (**41.4%**). This suggests no significant disparity.
:::

```{r, echo=FALSE, fig.width=8, fig.height=6, fig.align="center"}
# Compute percentages
HbA1c_counts <- HbA1c_by_gender %>%
  count(gender, HbA1c_category) %>%
  group_by(gender) %>%
  mutate(percent = n / sum(n) * 100)  # Compute percentage

# Create stacked bar plot with percentage labels
HbA1c_counts_plotly <- ggplot(HbA1c_counts, aes(x = gender, y = n, fill = HbA1c_category)) +
  geom_bar(stat = "identity", position = "fill", color = "black", linewidth = 0.1, alpha = 0.5) +
  geom_text(aes(label = paste0(round(percent, 1), "%")), 
            position = position_fill(vjust = 0.5)) +
  scale_fill_manual(values = c("Normal < 5.7%" = "cornsilk", 
                               "Prediabetes 5.7% - 6.4%" = "darkkhaki", 
                               "Diabetes ≥ 6.5%" = "darkgoldenrod4")) +
  labs(title = "Male vs. Female Blood Sugar Levels (HbA1c)", 
       x = "Gender", y = "Proportion", fill = "HbA1c Category") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))+
  theme(
    plot.title = element_text(size = 13),
    axis.text = element_text(size = 11),
  )

ggplotly(HbA1c_counts_plotly)
```

### BMI Distribution by Hypertension Status Plot

Shows the distribution of BMI values based on hypertension status. A violin plot is great for visualizing the distribution and density of BMI across hypertension categories,

Shape and width: The width of each "violin" represents the density of BMI values at different levels. Wider sections mean more individuals have that BMI, while narrower sections indicate fewer people at those values.

Comparison of distributions: The blue violin represents people without hypertension (hypertension = 0), while the red violin represents those with hypertension (hypertension = 1). By comparing them, you can see how BMI differs between these groups.

The horizontal line around 25 BMI: This marks the median BMI for each group. Since both violins have a horizontal line in roughly the same position, it suggests that the median BMI is around 25 for both hypertensive and non-hypertensive individuals.

Density trends: If the violins have different thicknesses in certain BMI ranges, it tells you which BMI values are more or less common in each group. People with hypertension seem to have a higher BMI overall, but both groups share a similar median.

The distribution shape is different—for example, if one violin is wider at higher BMI values, it suggests that hypertension is more common among individuals with higher BMI.

Outliers or extreme values might appear as small bulges or extended tails at the ends of the violins, showing individuals with very high or low BMI.

```{r, echo=FALSE,fig.width=7, fig.height=6, fig.align="center"}
library(dplyr)
library(ggplot2)
library(plotly)

overall_median_bmi <- median(diabetes_dataset$bmi, na.rm = TRUE)

BMI_distribution <- ggplot(diabetes_dataset, aes(x = factor(hypertension), y = bmi, fill = factor(hypertension))) +
  geom_violin(alpha = 0.6) +
  geom_hline(yintercept = overall_median_bmi, linetype = "solid", color = "red", linewidth = 0.3) +
  scale_fill_manual(values = c("azure3", "#CC6600")) +
   scale_y_continuous(breaks = seq(10.01, 95.69, by = 5)) +
  labs(title = "BMI Distribution by Hypertension Status",
       x = "Hypertension Status (0 = No, 1 = Yes)",
       y = "BMI",
       fill = "Hypertension") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))+
  theme(
    plot.title = element_text(size = 13),
    axis.text = element_text(size = 11),
  )

ggplotly(BMI_distribution)
```

The graph below is separated by whether or not a person has hypertension. With the comparison of BMI as the range, it's seen that majority of people with and without hypertension lie within a BMI range of 25-29. Notice that for people with hypertension, the desnity population above the red line is greater than that of people without hypertension; indicating that there's a larger of population of people with hypertension that have a larger BMI  



```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=6}
library(dplyr)
library(ggplot2)
library(plotly)

BGL_status_diabetes_dataset <- diabetes_dataset %>%
  select(blood_glucose_level,diabetes, age) %>%
  filter(age >= 3) %>% 
  mutate(diabetes = factor(diabetes, levels = c(0, 1), labels = c("No Diabetes", "Diabetes"))) %>% 
  ggplot(aes(x = diabetes, y = blood_glucose_level, fill = diabetes)) +
  geom_boxplot(alpha = 0.5) +
  labs(title = "Blood Glucose Levels by Diabetes Status (age 3-80)",
       x = "Diabetes Status",
       y = "Blood Glucose Level") +
  scale_fill_manual(values = c("Diabetes" = "#CC3300", "No Diabetes" = "#99CCFF")) +
  labs(fill = "Status") +
  scale_y_continuous(breaks = seq(80, 300, by = 20)) +
  geom_hline(yintercept = c(120, 140), linetype = "solid", color = "#0033FF", size = 0.2) +
  geom_hline(yintercept = c(180, 250), linetype = "solid", color = "#CC3300", size = 0.3) +  # Mark High Range
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))+
  theme(
    plot.title = element_text(size = 13),
    axis.text = element_text(size = 11),
  )

ggplotly(BGL_status_diabetes_dataset)
```

## BMI vs. Age Across Diabetes & Heart Disease Plot

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=6, fig.align="center"}
library(ggplot2)
library(dplyr)

diabetes_dataset_filtered_bmi <- diabetes_dataset %>%
  mutate(bmi_filtered = ifelse(bmi >= 2, bmi, NA))

diabetes_only <- diabetes_dataset_filtered_bmi %>%
  select(age, bmi_filtered, diabetes) %>%
  filter(age >= 2, diabetes == 1) %>%
  mutate(condition = "Diabetes Only")

heart_disease_only <- diabetes_dataset_filtered_bmi %>%
  select(age, bmi_filtered, heart_disease) %>%
  filter(age >= 2, heart_disease == 1) %>%
  mutate(condition = "Heart Disease Only")

combined_data <- bind_rows(diabetes_only, heart_disease_only)

common_bmi <- diabetes_dataset_filtered_bmi %>%
  filter(bmi_filtered >= 25.01 & bmi_filtered <= 30.01) %>%
  count(bmi_filtered) %>%
  arrange(desc(n))  # Sort by highest count

most_common_bmi <- common_bmi$bmi_filtered[1]  # Extract the most frequent BMI

ggplot(combined_data) +
  geom_point(aes(x = age, y = bmi_filtered, color = condition), alpha = 0.3) +  # Scatter plot with BMI
  geom_jitter(aes(x = age, y = bmi_filtered, color = condition), width = 0.1, height = 0.1, alpha = 0.3) +
  geom_hline(yintercept = most_common_bmi, linetype = "solid", color = "red", size = 0.3) +  # Horizontal line marking common BMI
  scale_color_manual(values = c("Diabetes Only" = "cornflowerblue", "Heart Disease Only" = "darkorchid4")) + 
  scale_x_continuous(breaks = seq(0, 80, by = 5)) + 
  scale_y_continuous(breaks = seq(10.01, 95.69, by = 10)) +
  labs(title = "BMI vs. Age Across Diabetes & Heart Disease",
       x = "Age",
       y = "BMI",
       color = "Condition") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(
    axis.text = element_text(size = 11),
    legend.text = element_text(size = 11),
    plot.title = element_text(size = 13)
  )

```

# Smokers go brrr

In the smoking data there are 6 unique values

1)  Never: Has Never smoked
2)  Not current: Has smoked but is not currently smoking
3)  Former: Has quit smoking (abstained for longer than)
4)  Current: Is currently a smoker
5)  Ever: Has ever smoked regardless of current smoking status
6)  No Info: No smoking history information available

The total amount of people who fall into each category is as follows;

1)  Never: 35095
2)  Not current: 6447
3)  Former: 9352
4)  Current: 9286
5)  Ever: 4004
6)  No Info: 35816

There is quite a sizable amount of people in the No info category.

The total number of people in the dataset is `r nrow(diabetes_dataset)`. To help clean up the data, we can filter 'No Info' people out. When we do that we get `r nrow(diabetes_dataset %>% filter(smoking_history != 'No Info'))`.

```{r smoking data, echo=FALSE}

# Figure out the unique categories of smoking history 
unique(diabetes_dataset$smoking_history)

# Count amount of people who belong to each unique category
diabetes_dataset %>% group_by(smoking_history) %>% summarise(total_people = n())

smoking_diabetes_dataset <- diabetes_dataset %>% 
  filter(smoking_history != 'No Info') %>%  
  group_by(smoking_history, diabetes) %>%  
  summarise(total = n())
```

Now we can graph the relationship between

```{r, echo=FALSE}

library(dplyr)
library(ggplot2)

# Data prep
bmi_heart_gender <- diabetes_dataset %>%
  filter(heart_disease == 1) %>%
  mutate(
    bmi_category = case_when(
      bmi <= 19 ~ "Underweight",
      bmi >= 30 ~ "Overweight",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(bmi_category)) %>%
  group_by(gender, bmi_category) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(gender) %>%
  mutate(percentage = round(100 * count / sum(count), 1))

# Make sure the bmi_category is an ordered factor so the line connects correctly
bmi_heart_gender$bmi_category <- factor(bmi_heart_gender$bmi_category,
                                        levels = c("Underweight", "Overweight"))

# Dot + Line plot
ggplot(bmi_heart_gender, aes(x = bmi_category, y = count, group = gender, color = gender)) +
  geom_line(aes(group = gender), position = position_dodge(width = 0.5), linewidth = 1) +
  geom_point(size = 5, position = position_dodge(width = 0.5)) +
  geom_text(aes(label = paste0(count, " (", percentage, "%)")),
            vjust = -1,
            position = position_dodge(width = 0.5)) +
  labs(
    title = "Heart Disease Cases by Gender and BMI Category (Dot-Line Plot)",
    x = "BMI Category",
    y = "Count of People with Heart Disease",
    color = "Gender"
  ) +
  theme_minimal()


```

```{r, echo=FALSE}
library(dplyr)
library(ggplot2)

# Prepare the data
bmi_heart_gender <- diabetes_dataset %>%
  filter(heart_disease == 1) %>%
  mutate(
    bmi_category = case_when(
      bmi <= 19 ~ "Underweight",
      bmi >= 30 ~ "Overweight",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(bmi_category)) %>%
  group_by(gender, bmi_category) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(gender) %>%
  mutate(percentage = round(100 * count / sum(count), 1))

# Plot
ggplot(bmi_heart_gender, aes(x = bmi_category, y = count, fill = gender)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.7)) +
  geom_text(aes(label = paste0(count, " (", percentage, "%)")),
            vjust = -0.5,
            position = position_dodge(width = 0.7)) +
  labs(
    title = "Count and Percentage of Heart Disease Cases by Gender and BMI Category",
    x = "BMI Category",
    y = "Count of People with Heart Disease",
    fill = "Gender"
  ) +
  theme_minimal()

```

```{r, echo=FALSE}
# load needed packages
library(dplyr)
library(ggplot2)

 male_data = diabetes_dataset %>% filter(gender == "Male")
 female_data = diabetes_dataset %>% filter(gender == "Female")
# 1. Summarise men
 
men_summary <- male_data %>%
  summarise(
    count = sum(bmi >= 30 & diabetes == 0),
    total = n()
  ) %>%
  mutate(
    sex        = "Men",
    percent    = count / total * 100
  )

# 2. Summarise women
women_summary <- female_data %>%
  summarise(
    count = sum(bmi >= 30 & diabetes == 0),
    total = n()
  ) %>%
  mutate(
    sex        = "Women",
    percent    = count / total * 100
  )

# 3. Combine into one data frame
summary_df <- bind_rows(men_summary, women_summary)

# 4. Plot
ggplot(summary_df, aes(x = sex, y = count, fill = sex)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%d (%.1f%%)", count, percent)),
            vjust = -0.5, size = 4) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Count of Overweight (BMI ≥ 30) without diabetes, by Sex",
    x     = NULL,
    y     = "Number of Individuals",
    fill  = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

```


```{r}
library(dplyr)
library(ggplot2)
library(scales)

df_summary <- diabetes_dataset %>%
  mutate(
    HbA1c_cat = case_when(
      HbA1c_level < 5.7                      ~ "< 5.7 (non‑diabetic)",
      HbA1c_level >= 5.7 & HbA1c_level < 6.5 ~ "5.7–6.4 (prediabetic)",
      HbA1c_level >= 6.5                    ~ "≥ 6.5 (diabetic)"
    ),
    hypertension = factor(hypertension, levels = c(0,1),
                          labels = c("No", "Yes"))
  ) %>%
  group_by(HbA1c_cat, hypertension) %>%
  summarise(n = n(), .groups = "drop_last") %>%
  mutate(percent = n / sum(n))  # auto‑groups by HbA1c_cat




ggplot(df_summary, aes(x = HbA1c_cat,
                       y = percent,
                       fill = factor(hypertension))) +
  geom_col(position = "stack") +
  geom_text(aes(label = scales::percent(percent, 1)),
            position = position_stack(vjust = 0.5)) +
  scale_y_continuous(labels = scales::percent_format()) +
  scale_fill_discrete(name = "Hypertension",
                      labels = c("No", "Yes")) +
  labs(
    title = "Hypertension Status by HbA1c Category",
    x     = "HbA1c Category",
    y     = "Percent within Category"
  ) +
  theme_light()

```








# Smokers go brrr# Smokers go brrrgeom_boxplot()

In the smoking data there are 6 unique values

1)  Never: Has Never smoked
2)  Not current: Has smoked but is not currently smoking
3)  Former: Has quit smoking (abstained for longer than)
4)  Current: Is currently a smoker
5)  Ever: Has ever smoked regardless of current smoking status
6)  No Info: No smoking history information available

The total amount of people who fall into each category is as follows;

1)  Never: 35095
2)  Not current: 6447
3)  Former: 9352
4)  Current: 9286
5)  Ever: 4004
6)  No Info: 35816

There is quite a sizable amount of people in the No info category.

The total number of people in the dataset is `r nrow(diabetes_dataset)`. To help clean up the data, we can filter 'No Info' people out. When we do that we get `r nrow(diabetes_dataset %>% filter(smoking_history != 'No Info'))`.

```{r Cleaning smoking data}

# Figure out the unique categories of smoking history
unique(diabetes_dataset$smoking_history)

# Count amount of people who belong to each unique smoking category
#Omit No info

smoking_tally <- diabetes_dataset %>% filter(smoking_history != 'No Info') %>%  group_by(smoking_history) %>% summarise(total_people = n())

#group diabetic vs non diabetic people together

smoking_diabetes_dataset <- diabetes_dataset %>%
  filter(smoking_history != 'No Info') %>%  
  group_by(smoking_history, diabetes) %>%  
  summarise(total = n())

smoking_diabetes_dataset

# Inner join tally data with diabetic grouped data;
#mutate a column to create a percentage per category;
#select desired columns

smoking_diabetes_percentage <- inner_join(smoking_tally, smoking_diabetes_dataset, by = 'smoking_history') %>% mutate(Percentage = total/total_people *100) %>% select(smoking_history, diabetes, total, Percentage)

smoking_diabetes_percentage
```

Now we can graph the relationship between smoking and diabetes as separated by smoking category.

```{r Plotting smoking data}
library(ggplot2)
library(plotly)

#Create initial graph about smoking and diabetes

smoking_diabetes_graph <- ggplot(smoking_diabetes_percentage) + 
  geom_col(aes(x = smoking_history, y = total, fill = as.factor(diabetes)), position = 'dodge') 

smoking_diabetes_graph <- smoking_diabetes_graph + coord_flip() + labs(
  title = 'Smoking History and Diabetes Relationship',
  y = 'People',
  x = 'Smoking History',
  fill = 'Has Diabetes',
  caption = "1 indicates diabetes, 0 indicates no diabetes"
  ) 

smoking_diabetes_graph
```


```{r Smoking on that percentage pack tonight tonight}

ggplot(smoking_diabetes_percentage) + geom_col(aes(x = smoking_history, y = Percentage, fill = as.factor(diabetes)))
```